---
title: 'Survival Analysis'
teaching: 10
exercises: 2
---

```{r, include=FALSE}
library(magrittr)
library(tidyverse)
library(grafify)
library(cowplot)
library(brms)
library(bayesplot)
library(splines2)
theme_set(theme_bw(15))


okabi_colors <- c("#E69F00", "#56B4E9", "#009E73")

prior_color <- "#009E73"
likelihood_color <- "#E69F00"
posterior_color <- "#56B4E9"

bayesplot::color_scheme_set("brightblue")
```

::: questions
-   How does Bayesian modeling accommodate survival analysis?
:::

::: objectives
-   Understand survival analysis basics: hazard, survival function, censoring

-   Fit and interpret Weibull (parametric) and Cox (semiparametric) survival models using Stan

-   Explore more advanced model that has hierarchical structure
:::

Survival (time-to-event) data arise when the outcome of interest is the time until an event occurs, for example device failure or death. Unlike ordinary regression, survival data often include censoring, where the exact event time is not fully observed for some units. Analyses must therefore use models that account for both observed event times and censored observations. A simple descriptive summary is the KaplanŌĆōMeier curve, which estimates the survival function $$
S(t) = P(T > t)
$$and handles right censoring naturally. We will use KaplanŌĆōMeier estimates for exploratory checks alongside model-based summaries.

Two functions organize survival modeling. The first is the survival function $S(t) = P(T > t)$. The second is the hazard function $h(t)$, risk for an event at time t among those still at risk. Formally, $$h(t) = \lim_{dt \to 0} \dfrac{P(t \le T < t + dt \mid T \ge t)}{dt}.
$$It is often useful to consider the cumulative hazard $H(t) = \int_0^t h(u)\,du$, which links back to survival through $S(t) = \exp\{-H(t)\}$.

Censoring is part of the data generating process in many applications. Right censoring occurs when we only know that the event time exceeds the observed follow-up. Left censoring occurs when we only know that the event happened before a detection or assessment time. Left truncation, also called delayed entry, occurs when a unit enters the risk set after time zero. Correct handling of the risk set and the likelihood contributions for these cases is essential for valid inference.

There are several modeling strategies for time-to-event outcomes. In a parametric approach such as the Weibull model, we specify a distribution for event times, which yields interpretable parameters and straightforward extrapolation. In a semiparametric approach such as the Cox proportional hazards model, we leave the baseline hazard unspecified and model covariate effects multiplicatively, $h(t \mid x) = h_0(t)\,\exp(x^\top \beta)$. Parametric models can be more efficient when the assumed form is appropriate. Cox models offer robustness when the baseline hazard shape is unknown.

This episode adopts a Bayesian perspective consistent with the rest of the course. Censoring enters the likelihood directly. Observed events contribute the density $f(t \mid x)$ and right censored observations contribute the survival $S(t \mid x)$. Priors provide regularization, which can be especially helpful when the number of events is small or when models include many parameters. Posterior draws allow direct uncertainty quantification for quantities of interest such as hazard ratios $\exp(\beta)$, median survival times, survival curves at selected covariate values, and contrasts between treatments or groups.

We will use Stan directly from R to implement a few survival models. Specifically, we will code and fit a parametric Weibull model and a Cox proportional hazards model for censored time-to-event data.

## Survival data

In this example we fit a Weibull survival model using only microbiome features as predictors. The dataset includes 150 mice and 48 microbial genera, with follow-up recorded as the time to type 1 diabetes onset or censoring. Half of the mice experienced the event (T1D onset) and half were censored.

All microbial covariates are standardized to have mean 0 and standard deviation 1 before modeling. Consequently, hazard ratios are interpreted per oneŌĆōstandard deviation increase in a covariate, not per one unit in the original measurement scale.

```{r, message=FALSE, warning=FALSE}
# Packages
library(dplyr)
library(rstan)
library(posterior)
library(survival)
library(TreeSummarizedExperiment)
library(mia)
library(ggplot2)
library(survminer)
```

```{r, warning=FALSE, message=FALSE}
# Load TSE
tse <-  readRDS(file.path("data", "survival_tse.rds"))

# Relative abundance + log transform
tse <- tse |>
  transformAssay(method = "relabundance", assay.type = "counts", name = "relabundance") |>
  transformAssay(method = "log", assay.type = "relabundance", pseudocount = 1e-6, name = "log_abund")

# Features (samples x microbes)
df <- as.data.frame(t(assay(tse, "log_abund")))

# Outcomes
df$Event      <- colData(tse)$Event
df$Event_time <- colData(tse)$Event_time

# Pick 10 random microbes as covariates
set.seed(2025)
feature_cols <- setdiff(names(df), c("Event", "Event_time"))
selected_microbes <- sample(feature_cols, min(10L, length(feature_cols)))

# Build design matrix (scaled)
X <- scale(as.matrix(df[, selected_microbes, drop = FALSE]))
K <- ncol(X)

# Analysis data frame
analysis_df <- data.frame(
  time  = as.numeric(df$Event_time),
  event = as.integer(df$Event),
  X,
  check.names = FALSE
)
rownames(analysis_df) <- rownames(df)

# Stan data
standata <- list(
  N    = nrow(analysis_df),
  K    = K,
  time = as.vector(analysis_df$time),
  event= as.integer(analysis_df$event),
  X    = as.matrix(analysis_df[, colnames(X), drop = FALSE])
)

```

## Weibull survival model

The Weibull distribution is a versatile parametric model for time-to-event data. The Weibull model is flexible and often a good first choice. Let $T$ be the event time. It is defined by a shape parameter $k>0$ and a scale parameter $\lambda>0$. In survival analysis the survival function is 
$$
S(t)=\exp\left[-\left(\frac{t}{\lambda}\right)^{k}\right]
$$ 
and the hazard is 
$$ 
h(t) = \frac{k}{\lambda}\left(\frac{t}{\lambda}\right)^{k-1}. 
$$
The shape controls how risk evolves over time: when $0<k<1$ the hazard decreases, when $k=1$ it is constant (the exponential case), and when $k>1$ it increases. The scale $\lambda$ sets the time axis, stretching survival times as it grows. A convenient summary is the median survival time 
$$ 
\operatorname{med}(T)=\lambda(\log 2)^{1/k}, 
$$
which increases with $\lambda$ and, for fixed $\lambda$, decreases as $k$ increases. Thanks to this ability to capture decreasing, constant, or increasing hazards with a single parameter, the Weibull distribution is a natural starting point for modeling event times.

### Example

Next, we illustrate the parametric Weibull survival model by estimating hazard ratios and comparing risk groups with KaplanŌĆōMeier curves. Below is the structure of the Stan program.

```{stan output.var="weibull_model"}
// Weibull survival model (shape k, scale ╬╗), simple AFT form
data {
  int<lower=1> N;                    // observations
  int<lower=0> K;                    // number of covariates
  vector<lower=0>[N] time;           // observed times t_i
  int<lower=0,upper=1> event[N];     // 1=event, 0=right-censored
  matrix[N, K] X;                    // covariates (N x K, K can be 0)
}

parameters {
  real<lower=0> k;                   // Weibull shape
  real beta0;                        // intercept for log ╬╗
  vector[K] beta;                    // coefficients for log ╬╗
}

transformed parameters {
  vector[N] log_lambda;              // log ╬╗_i
  vector<lower=0>[N] lambda;         // ╬╗_i
  if (K > 0) log_lambda = beta0 + X * beta;
  else       log_lambda = rep_vector(beta0, N);
  lambda = exp(log_lambda);
}

model {
  // weakly-informative priors
  k     ~ lognormal(0, 1);
  beta0 ~ normal(0, 1);
  if (K > 0) beta ~ normal(0, 1);

  // likelihood: events use f(t), censoring uses log S(t)
  for (i in 1:N) {
    if (event[i] == 1)
      target += weibull_lpdf(time[i] | k, lambda[i]);   // log f(t_i)
    else
      target += weibull_lccdf(time[i] | k, lambda[i]);  // log S(t_i)
  }
}

generated quantities {
  vector[N] t_median;                // median survival: ╬╗_i * (log 2)^(1/k)
  for (i in 1:N)
    t_median[i] = lambda[i] * pow(log(2), 1.0 / k);
}

```

In our Weibull setup the scale depends on covariates via the AFT link $\log \lambda_i=\beta_0+x_i^\top\beta$ (equivalently, $\lambda_i=\exp(\beta_0+x_i^\top\beta)$), and the hazard is $h_i(t)=\dfrac{k}{\lambda_i}\bigl(\tfrac{t}{\lambda_i}\bigr)^{k-1}$. If we increase a single covariate $x_j$ by one unit (holding the others fixed), the hazard scales by

$$
\mathrm{HR}_j \;=\; \frac{h\!\left(t \mid x_j + 1\right)}{h\!\left(t \mid x_j\right)} \;=\; \exp\!\left(-k\,\beta_j\right).
$$

Thus a positive $\beta_j$ yields $\mathrm{HR}_j<1$ (lower hazard; protective) and a negative $\beta_j$ yields $\mathrm{HR}_j>1$ (higher hazard). In the code, we obtain HRs by transforming posterior draws of $k$ and $\beta_j$ with $\exp(-k\,\beta_j)$, and we summarize them by the posterior median and the $95%$ credible interval (2.5%ŌĆō97.5%).

The next chunk transforms posterior draws of $k$ and $\beta_j$ into hazard-ratio draws via $\exp(-k\,\beta_j)$, summarizes each covariate by the posterior median and 95% credible interval, and then produces an ordered forest plot highlighting covariates whose intervals do not include 1. The model is fit with 4 chains and 4000 iterations per chain.

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit the Weibull model
fit <- rstan::sampling(
  object  = weibull_model,
  data    = standata,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

# Print summary
print(fit, pars = c("k","beta0","beta"))

# Posterior draws
M <- posterior::as_draws_matrix(fit)
k_draw <- M[, "k"]
K <- standata$K
covar_names <- colnames(standata$X)

# Weibull AFT hazard ratios
hr_mat <- do.call(rbind, lapply(seq_len(K), function(j) {
  bj <- M[, paste0("beta[", j, "]")]
  hr_draw <- exp(- k_draw * bj)
  c(
    med = median(hr_draw),
    lo  = unname(quantile(hr_draw, 0.025)),
    hi  = unname(quantile(hr_draw, 0.975))
  )
}))
hr_df <- as.data.frame(hr_mat)
hr_df$term <- covar_names
names(hr_df) <- c("HR","lower","upper","term")

# Significance
hr_df$signif <- (hr_df$lower > 1) | (hr_df$upper < 1)
hr_df <- hr_df[order(hr_df$HR), ]
hr_df$term <- factor(hr_df$term, levels = hr_df$term)

# Plot a forest plot of hazard ratios
ggplot(hr_df, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"), guide = "none") +
  labs(title = "Weibull hazard ratios",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

Hazard ratios left of 1 (with 95% intervals not crossing 1) suggest protective associations while those right of 1 indicate increased hazard. In this plot, several taxa fall clearly left of 1, while a few lie right of 1. Estimates whose intervals cross 1 are inconclusive.

To visualize how the Weibull model separates risk, we form a model-based risk score from the AFT linear predictor $\log\lambda_i=\beta_0+x_i^\top\beta$ using posterior medians of $\beta_0$ and $\beta$. We split subjects at the median of this score (Low vs High) and compare their KaplanŌĆōMeier curves with a log-rank test.

```{r, fig.width=10, fig.height=6, dpi=300, warning=FALSE}
# Posterior medians for the AFT linear predictor
beta0_med <- median(M[, "beta0"])
beta_med  <- sapply(seq_len(K), function(j) median(M[, paste0("beta[", j, "]")]))

# Model-based score
risk_lp <- as.numeric(beta0_med + as.matrix(standata$X) %*% beta_med)

# Data for KM
dat_km <- data.frame(
  time  = standata$time,
  event = standata$event,           # 1 = event, 0 = censored
  risk  = risk_lp
)

# Median split (High score = better survival)
cut_pt <- median(dat_km$risk, na.rm = TRUE)
dat_km$group <- factor(
  ifelse(dat_km$risk > cut_pt, "High score", "Low score"),
  levels = c("Low score", "High score")
)

# KaplanŌĆōMeier fit and plot
sf <- survfit(Surv(time, event) ~ group, data = dat_km)

ggsurvplot(
  sf, data = dat_km,
  conf.int    = TRUE,
  pval        = TRUE,
  legend.labs = levels(dat_km$group),
  palette     = c("#E64B35", "#4DBBD5"),
  xlab = "Time", ylab = "Survival probability"
)

```

KaplanŌĆōMeier curves by the model-based score (median split) show clear separation: the high AFT score group (larger $\log\lambda$, implying longer times) has higher survival (curve above), while the low-score group declines faster. The log-rank test indicates a statistically significant difference.



:::::::::::::::::::::::::::::: challenge

Pick two microbial covariates from the Weibull forest plot:  
one with HR < 1 and one with HR > 1.

1. Explain in words what each hazard ratio means for time to T1D onset.  
2. Remember that covariates are standardized: interpret the HRs per oneŌĆōstandard deviation increase.
 

::::::::::::::::::::::: solution

**HR < 1**  
A oneŌĆōstandard deviation increase in the microbial covariate is associated with a lower instantaneous risk of T1D onset (protective effect).

**HR > 1**  
A oneŌĆōstandard deviation increase in the microbial covariate is associated with a higher instantaneous risk of T1D onset (harmful effect).

::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::



## Cox proportional hazards model

The Cox model is a semiparametric approach for time-to-event data in which covariates act multiplicatively on the hazard. Let $T$ be the event time and $x$ the covariates. The hazard is

$$
h(t \mid x) \;=\; h_0(t)\,\exp\!\bigl(x^\top \beta\bigr),
$$

where $h_0(t)$ is an unspecified baseline hazard. Through the cumulative baseline hazard $H_0(t)=\int_0^t h_0(u)\,du$, the survival function can be written

$$
S(t \mid x) \;=\; \exp\!\Bigl[-\,H_0(t)\,\exp\!\bigl(x^\top \beta\bigr)\Bigr]
\;=\; S_0(t)^{\,\exp(x^\top \beta)}.
$$

The coefficients $\beta$ are logŌĆōhazard ratios: a one-unit increase in $x_j$ multiplies the hazard by $\exp(\beta_j)$. Estimation is based on the partial likelihood, which compares each eventŌĆÖs linear predictor to those of individuals still at risk at that time and avoids modeling $h_0(t)$. The key assumption is proportional hazards, meaning the hazard ratio for two covariate values is constant over time.

### Example

Next, we illustrate the Cox proportional hazards model by estimating hazard ratios and comparing risk groups with KaplanŌĆōMeier curves. We fit the model via the partial likelihood (leaving the baseline hazard unspecified) and then summarize covariate effects as exponentiated coefficients $exp(\beta)$. Below is the structure of the Stan program.

```{stan output.var="cox_model"}
// Cox proportional hazards model, partial likelihood (PH form)
data {
  int<lower=1> N;                    // observations
  int<lower=0> K;                    // number of covariates
  vector<lower=0>[N] time;           // observed times t_i
  int<lower=0, upper=1> event[N];    // 1=event, 0=right-censored
  matrix[N, K] X;                    // covariates (N x K, K can be 0)
}

transformed data {
  // Indices of observed events 
  int<lower=0> n_event = 0;
  int event_idx[N];
  for (i in 1:N) {
    if (event[i] == 1) {
      n_event += 1;
      event_idx[n_event] = i;
    }
  }
}

parameters {
  vector[K] beta;                    // log hazard ratios
}

model {
  // Weakly-informative prior
  if (K > 0) beta ~ normal(0, 1);

  // Cox partial log-likelihood
  vector[N] eta;
  if (K > 0) eta = X * beta; else eta = rep_vector(0, N);

  for (e in 1:n_event) {
    int i = event_idx[e];
    real lse = negative_infinity();   // log(0)
    for (j in 1:N) {
      if (time[j] >= time[i])         // j in risk set at t_i
      lse = log_sum_exp(lse, eta[j]);
    }
    target += eta[i] - lse;
  }
}

generated quantities {
  vector[K] HR;                        // per-covariate hazard ratios
  for (k in 1:K) HR[k] = exp(beta[k]);
}

```

In our Cox proportional hazards setup the hazard for subject $i$ is modeled as $$
h_i(t) = h_0(t)\,\exp\!\big(x_i^\top\beta\big),
$$ where $h_0(t)$ is an unspecified baseline hazard and $\beta$ are logŌĆōhazard ratio coefficients. If we increase a single covariate $x_j$ by one unit (holding the others fixed), the hazard scales by $$
\mathrm{HR}_j \;=\; \frac{h\!\left(t \mid x_j + 1\right)}{h\!\left(t \mid x_j\right)} \;=\; \exp(\beta_j).
$$ Unlike the Weibull AFT parameterization (where $\mathrm{HR}_j=\exp(-k\,\beta_j)$ so a positive $\beta_j$ is protective), in the Cox PH model a positive $\beta_j$ yields $\mathrm{HR}_j>1$ (higher hazard) and a negative $\beta_j$ yields $\mathrm{HR}_j<1$ (lower hazard is protective). In the code, we obtain HRs directly from posterior draws of $\beta_j$ via $\exp(\beta_j)$ and summarize them by the posterior median and the $95\%$ credible interval (2.5%ŌĆō97.5%).

The next chunk transforms posterior draws of $\beta_j$ into hazard-ratio draws via $\exp(\beta_j)$, summarizes each covariate by the posterior median and 95% credible interval, and then produces an ordered forest plot highlighting covariates whose intervals do not include 1. The model is fit with 4 chains and 4000 iterations per chain.

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit the Cox model
fit <- rstan::sampling(
  object  = cox_model,
  data    = standata,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

# Print summary
print(fit, pars = c("beta"))

# Posterior draws
M <- posterior::as_draws_matrix(fit)
K <- standata$K
covar_names <- colnames(standata$X)

# Cox PH hazard ratios
hr_mat <- do.call(rbind, lapply(seq_len(K), function(j) {
  bj <- M[, paste0("beta[", j, "]")]
  hr_draw <- exp(bj)
  c(
    med = median(hr_draw),
    lo  = unname(quantile(hr_draw, 0.025)),
    hi  = unname(quantile(hr_draw, 0.975))
  )
}))
hr_df <- as.data.frame(hr_mat)
hr_df$term <- covar_names
names(hr_df) <- c("HR","lower","upper","term")

# Significance
hr_df$signif <- (hr_df$lower > 1) | (hr_df$upper < 1)
hr_df <- hr_df[order(hr_df$HR), ]
hr_df$term <- factor(hr_df$term, levels = hr_df$term)

# Plot a forest plot of hazard ratios
ggplot(hr_df, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"), guide = "none") +
  labs(title = "Cox PH hazard ratios",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

For the Cox PH model, the hazardŌĆÉratio pattern mirrors the Weibull results: hazard ratios left of 1 (with 95% intervals not crossing 1) suggest protective associations, while those right of 1 indicate increased hazard; intervals that cross 1 are inconclusive. In our analysis, the Cox HRs align closely with the Weibull HRs, leading to very similar forest plots.

To visualize how the Cox model separates risk, we form a model-based risk score from the PH linear predictor $\eta_i = x_i^\top \beta$ using posterior medians of $\beta$. We split subjects at the median of this score (ŌĆ£LowŌĆØ vs ŌĆ£HighŌĆØ) and compare their KaplanŌĆōMeier curves with a log-rank test. The KaplanŌĆōMeier estimator is nonparametric and does not depend on the Cox model itself, only the grouping into ŌĆ£LowŌĆØ and ŌĆ£HighŌĆØ risk comes from the model, so the same KM machinery could equally be used with the Weibull-based risk score.

```{r, fig.width=10, fig.height=6, dpi=300, warning=FALSE}
# Posterior medians for the PH linear predictor
beta_med <- sapply(seq_len(K), function(j) median(M[, paste0("beta[", j, "]")]))
risk_lp  <- as.numeric(as.matrix(standata$X) %*% beta_med)

dat_km <- data.frame(
  time  = standata$time,
  event = standata$event,
  risk  = risk_lp
)

# Median split of the risk score (High risk = worse survival)
cut_pt <- median(dat_km$risk, na.rm = TRUE)
dat_km$group <- factor(
  ifelse(dat_km$risk > cut_pt, "High risk", "Low risk"),
  levels = c("High risk", "Low risk")
)

# KaplanŌĆōMeier fit and plot
sf <- survfit(Surv(time, event) ~ group, data = dat_km)

ggsurvplot(
  sf, data = dat_km,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Model-based risk",
  legend.labs  = levels(dat_km$group),
  palette      = c("#E64B35", "#4DBBD5"),
  xlab         = "Time", ylab = "Survival probability"
)


```

KaplanŌĆōMeier curves by the model-based score (median split) show clear separation: the high PH risk group (larger $\eta = x^\top\beta$, implying higher hazard) has lower survival (curve below), while the low-risk group remains higher. The log-rank test indicates a statistically significant difference.

## Hierarchical survival model

In the final example we switch from the microbiome data to a clinical survival dataset. The cancer data from the survival package contains follow-up on patients with advanced lung cancer, with time recorded from study entry to death or censoring. Each patient has several baseline covariates, including age, sex, ECOG performance status, physician- and patient-rated Karnofsky scores, daily caloric intake, and weight loss in last six months. In addition, patients are treated at different institutions, indexed by the variable inst.

To capture both individual-level risk factors and between-institution heterogeneity, we construct a hierarchical Cox proportional hazards model. All baseline covariates enter the linear predictor as fixed effects, while the institution index defines group-specific frailties that are modeled as random effects on the log-hazard scale. For patient $i$ treated at institution $g[i]$, the hazard is

$$
h_i(t) \;=\; h_0(t)\,\exp\bigl(x_i^\top \beta + u_{g[i]}\bigr), 
\qquad u_g \sim \mathcal{N}(0, \sigma_u^2),
$$

where $x_i$ is the vector of scaled baseline covariates, $\beta$ are fixed-effect coefficients, and $u_g$ is the institution-specific frailty term with variance $\sigma_u^2$. All baseline covariates are standardized (mean 0, standard deviation 1), so hazard ratios summarize the effect of a oneŌĆōstandard deviation increase in each covariate.

```{r}
# Data
data(cancer, package = "survival")

# Complete cases
cancer2 <- na.omit(cancer)

# Outcome: time and event indicator (status: 1=censored, 2=dead)
cancer2$time  <- cancer2$time
cancer2$event <- as.integer(cancer2$status == 2)

# Grouping factor: institution
cancer2$group <- factor(cancer2$inst)
G <- nlevels(cancer2$group)

# All baseline covariates
X_mm <- model.matrix(
  ~ age + sex + ph.ecog + ph.karno + pat.karno + meal.cal + wt.loss,
  data = cancer2
)[, -1, drop = FALSE]  # drop intercept

# Scale columns
X <- scale(X_mm)
K <- ncol(X)

# Analysis data frame
analysis_df_cancer <- data.frame(
  time  = cancer2$time,
  event = cancer2$event,
  group = cancer2$group,
  X,
  check.names = FALSE
)

# Stan data list for hierarchical Cox PH
standata_cancer <- list(
  N     = nrow(analysis_df_cancer),
  G     = G,
  group = as.integer(analysis_df_cancer$group),
  K     = K,
  time  = as.vector(analysis_df_cancer$time),
  event = as.integer(analysis_df_cancer$event),
  X     = as.matrix(analysis_df_cancer[, colnames(X), drop = FALSE])
)

```

Next we implement this hierarchical specification in Stan by extending the Cox proportional hazards model with a random intercept for each institution. The linear predictor combines fixed effects for the scaled covariates with a group-specific frailty term, and inference proceeds again via the partial likelihood. As before, covariate effects are expressed as hazard ratios obtained by exponentiating the regression coefficients, allowing us to quantify how each predictor multiplies the event hazard while the frailty term accounts for unexplained heterogeneity between institutions.

```{stan output.var="cox_hier_model"}
// Stan: hierarchical Cox proportional hazards model
data {
  int<lower=1> N;                     // observations
  int<lower=1> G;                     // groups
  int<lower=1,upper=G> group[N];      // group index
  int<lower=0> K;                     // number of covariates
  vector<lower=0>[N] time;            // observed times t_i
  int<lower=0,upper=1> event[N];      // 1=event, 0=censored
  matrix[N, K] X;                     // covariates
}

transformed data {
  int<lower=0> n_event = 0;
  int event_idx[N];
  for (i in 1:N) {
    if (event[i] == 1) {
      n_event += 1;
      event_idx[n_event] = i;
    }
  }
}

parameters {
  vector[K] beta;                     // log hazard ratios
  vector[G] u_raw;                    // group effects
  real<lower=0> sigma_u;              // group SD
}

model {
  vector[N] eta;

  if (K > 0) beta ~ normal(0, 1);
  u_raw   ~ normal(0, 1);
  sigma_u ~ normal(0, 1);

  for (i in 1:N)
    eta[i] = (K > 0 ? X[i] * beta : 0) + sigma_u * u_raw[group[i]];

  for (e in 1:n_event) {
    int i = event_idx[e];
    real lse = negative_infinity();
    for (j in 1:N) {
      if (time[j] >= time[i])
        lse = log_sum_exp(lse, eta[j]);
    }
    target += eta[i] - lse;
  }
}

generated quantities {
  vector[K] HR;                       // hazard ratios
  real var_frailty = square(sigma_u); // between-group variance

  for (k in 1:K)
    HR[k] = exp(beta[k]);
}


```

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit hierarchical Cox and plot Hazard Ratios
fit_cancer_cox_hier <- rstan::sampling(
  object  = cox_hier_model,
  data    = standata_cancer,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

# Summary: fixed effects + frailty variance
print(fit_cancer_cox_hier, pars = c("beta","sigma_u","var_frailty","HR"))

# Posterior draws
M_cancer_cox <- posterior::as_draws_matrix(fit_cancer_cox_hier)
K <- standata_cancer$K
covar_names_cancer <- colnames(standata_cancer$X)

# Cox PH hazard ratios
hr_df_cancer <- do.call(rbind, lapply(seq_len(K), function(j) {
  hr_draw <- exp(M_cancer_cox[, paste0("beta[", j, "]")])
  data.frame(
    HR    = median(hr_draw),
    lower = unname(quantile(hr_draw, 0.025)),
    upper = unname(quantile(hr_draw, 0.975)),
    term  = covar_names_cancer[j],
    stringsAsFactors = FALSE
  )
}))

hr_df_cancer$signif <- (hr_df_cancer$lower > 1) | (hr_df_cancer$upper < 1)
hr_df_cancer <- hr_df_cancer[order(hr_df_cancer$HR), , drop = FALSE]
hr_df_cancer$term <- factor(hr_df_cancer$term, levels = hr_df_cancer$term)

ggplot(hr_df_cancer, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"),
                     guide = "none") +
  labs(title = "Hierarchical Cox PH hazard ratios (cancer data)",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

Performance status variables (ph.ecog and ph.karno) have hazard ratios above 1, with intervals not overlapping 1, indicating strong evidence that poorer performance status is associated with higher mortality. Age also shows a harmful effect, though with greater posterior uncertainty. In contrast, patient-rated Karnofsky score and caloric intake (meal.cal) have hazard ratios centered below or near 1, suggesting weak or mild protective associations that are mostly uncertain. Weight loss shows a hazard ratio slightly below/around 1, but with wide uncertainty, so its association with survival is unclear. Sex (female lower hazard) appears protective with a hazard ratio below 1.

The frailty variance is small but non-zero, indicating modest but detectable between-institution heterogeneity: some institutions have slightly higher or lower baseline hazard even after adjusting for patient-level predictors, but these differences are limited in magnitude compared to fixed effects.

Overall, performance status dominates the risk profile, while other baseline variables show smaller and more uncertain associations. The hierarchical model accounts for institutional heterogeneity but the main structure of covariate effects remains driven by individual-level predictors.

Next we split subjects at the median of this score (ŌĆ£LowŌĆØ vs ŌĆ£HighŌĆØ) and compare their KaplanŌĆōMeier curves with a log-rank test.

```{r, fig.width=10, fig.height=6, dpi=300}
# Posterior medians for the PH linear predictor
beta_med_cancer <- sapply(seq_len(K), function(j) {
  median(M_cancer_cox[, paste0("beta[", j, "]")])
})

# Model-based risk score (higher = worse survival)
risk_lp_cancer <- as.numeric(as.matrix(standata_cancer$X) %*% beta_med_cancer)

dat_km_cancer <- data.frame(
  time  = standata_cancer$time,
  event = standata_cancer$event,
  risk  = risk_lp_cancer
)

# Median split of the risk score (High risk = worse survival)
cut_pt_cancer <- median(dat_km_cancer$risk, na.rm = TRUE)
dat_km_cancer$group <- factor(
  ifelse(dat_km_cancer$risk > cut_pt_cancer, "High risk", "Low risk"),
  levels = c("High risk", "Low risk")
)

# KaplanŌĆōMeier fit and plot
sf_cancer <- survfit(Surv(time, event) ~ group, data = dat_km_cancer)

ggsurvplot(
  sf_cancer, data = dat_km_cancer,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Model-based risk",
  legend.labs  = levels(dat_km_cancer$group),
  palette      = c("#E64B35", "#4DBBD5"),
  xlab         = "Time", ylab = "Survival probability"
)

```

KaplanŌĆōMeier curves by the model-based score (median split) show clear separation: the high PH risk group has lower survival, while the low-risk group remains higher. The log-rank test indicates a statistically significant difference.

::: keypoints
-   Survival analysis models time until an event while handling censoring and aims to describe and compare time-to-event patterns across covariates or groups.

-   Time-to-event outcomes can be modeled parametrically with a Weibull model that specifies survival and hazard functions or semiparametrically with a Cox proportional hazards model that leaves the baseline hazard unspecified while estimating relative risk.

-   The hazard summarizes instantaneous event risk, hazard ratios compare risk between groups or covariate levels, and KaplanŌĆōMeier curves provide a nonparametric survival estimate often shown with a log-rank test.
:::



:::::::::::::::::::::::::::::: challenge

Fit a Bayesian Cox proportional hazards model with sex as the only covariate and then compare the model-based hazard ratio with nonparametric KaplanŌĆōMeier curves.

1. Fit the Cox PH model in Stan using the code above and inspect the posterior summary for `beta` and `HR`.  
2. Compute the posterior median hazard ratio (`HR_med`) and a 95% credible interval (`HR_ci`) for Female vs Male and interpret these in words.  
3. Plot KaplanŌĆōMeier curves for males and females and compare them to the Cox modelŌĆÖs posterior hazard ratio: do they tell a consistent story about differences in survival by sex?


::::::::::::::::::::::: solution

```{r}
# Data
data(cancer, package = "survival")
cancer2 <- na.omit(cancer)

# Outcome: time and event (status: 1=censored, 2=dead)
time  <- cancer2$time
event <- as.integer(cancer2$status == 2)

# Covariate: sex (0 = Male, 1 = Female)
sex_num <- ifelse(cancer2$sex == 2, 1, 0)

# X matrix
X_sex <- matrix(sex_num, ncol = 1)
colnames(X_sex) <- "sex_female"
K <- ncol(X_sex)

standata_sex <- list(
  N    = length(time),
  K    = K,
  time = as.vector(time),
  event= as.integer(event),
  X    = X_sex
)


# Fit Cox PH -model
fit_cox_sex <- rstan::sampling(
  object  = cox_model,      # same as in the lecture
  data    = standata_sex,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

```


```{r}
print(fit_cox_sex, pars = c("beta", "HR"))
```


```{r}
M_sex <- posterior::as_draws_matrix(fit_cox_sex)

# log-hazard ratio Female vs Male
beta_draw  <- M_sex[, "beta[1]"]
HR_draw    <- exp(beta_draw)

HR_med     <- median(HR_draw)
HR_ci      <- quantile(HR_draw, c(0.025, 0.975))
HR_med
HR_ci

```


```{r}
# Kaplan-Meier
cancer2$sex_factor <- factor(cancer2$sex, levels = c(1, 2),
                             labels = c("Male", "Female"))

sf_sex <- survfit(Surv(time, event) ~ sex_factor, data = cancer2)

ggsurvplot(
  sf_sex, data = cancer2,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Sex",
  legend.labs  = c("Male", "Female"),
  xlab         = "Time", ylab = "Survival probability"
)

```



1. Posterior median hazard ratio (Female vs Male) is about **0.63**. Females have roughly **37% lower** instantaneous risk of death than males at any given time.

2. The 95% credible interval is about **(0.43, 0.91)**, entirely below 1. This provides clear evidence that females have a lower hazard of death than males in this dataset.

3. The KaplanŌĆōMeier curve for females lies above the curve for males. Nonparametric KM estimates also indicate better survival for women, consistent with the Cox modelŌĆÖs hazard ratio.
   
::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::


## Resources

-   UserŌĆÖs guide <https://mc-stan.org/docs/stan-users-guide/survival.html>
