---
title: 'Survival Analysis'
teaching: 10
exercises: 2
---

```{r, include=FALSE}
library(magrittr)
library(tidyverse)
library(grafify)
library(cowplot)
library(brms)
library(bayesplot)
library(splines2)
theme_set(theme_bw(15))


okabi_colors <- c("#E69F00", "#56B4E9", "#009E73")

prior_color <- "#009E73"
likelihood_color <- "#E69F00"
posterior_color <- "#56B4E9"

bayesplot::color_scheme_set("brightblue")
```

::: questions
- What is survival analysis?   
- How to perform survival analysis with Stan?
:::

::: objectives
-   Understand survival analysis basics: hazard, survival function, censoring.

-   Learn to fit Weibull (parametric) and Cox proportional hazards (semiparametric) survival models using Stan.

-   Explore a more advanced model with hierarchical structure.
:::

Survival (or time-to-event) data arise when the outcome of interest is the time until some event occurs, such as device failure, disease onset or death. A key difference to ordinary regression settings is that time-to-event data are commonly censored, meaning that for some data points the event has not occurred by the end of the observation period. This implies that the exact event time is unknown and analyses must use techniques that account for both observed event times and censored observations.

:::::::::::::::::: callout 

In this episode, we will exclusively analyze right-censored data, meaning data where the event occurs during the follow-up time, or after this period. This is the most common type of censoring but there are two other censoring mechanisms are sometimes encountered. In *left censoring*, the event occurs before observation begins, so the event time is only known to be earlier than a given time point. *Interval censoring* means a situation where the event occurs between two observation times, but the exact timing is unknown. 

::::::::::::::::::

Before we start building survival models in Stan, we need to first get familiar with some mathematical concepts important in survival analysis. 

The two functions central to specifying common survival models are the survival function $S(t) = P(T > t)$ and the hazard function $h(t)$. The former gives the probability of surviving beyond time $t$, whereas the latter specifies the instantaneous risk for event at time $t$ among those who have not experienced an event yet. Formally the hazard is defined as $$h(t) = \lim_{dt \to 0} \dfrac{P(t \le T < t + dt \mid T \ge t)}{dt}.
$$ These two functions are linked through the identity $S(t) = \exp\{-H(t)\},$ where $H(t)$ is the cumulative hazard $H(t) = \int_0^t h(u)\,du$.

While a wealth of various tools for survival analysis  have been presented, including modern machine learning approaches, in this episode, we will focus on two relatively standard models: the parametric Weibull model and the semiparametric Cox proportional hazards model. Both of these model use the multiplicative form $h(t \mid x) = h_0(t)\,\exp(x^\top \beta),$ where  $h_0(t)$ is the baseline hazard describing hazard evolution when covariates equal zero.

A key distinction between the two models is that the Weibull model uses an explicit functional form for the baseline hazard (and simultaneously for the event times), while in the Cox model this is left unspecified. The Weibull models gains statistical strength from this assumed functional form. However, a mismatch between the assumption and data can lead to a poor model fit. The Cox model, on the other hand, can be more robust when the baseline hazard shape is unknown. Both approaches enable assessment of covariate effects of survival times, but only the Weibull model allows explicit survival time predictions.

The way censored events are accounted for in both models is by including the (incomplete) information in them in likelihood directly. In other words, the probability of event time being longer than the censoring threshold, namely $1 - F_T(t^{cens})$, where $F_T(t) = \text{Pr}(T < t)$ is the cumulative distribution function of survival time. Below, we will demonstrate how this is coded in Stan but first we will review our survival data. 

<!-- Present and demo these below -->
<!-- Posterior draws allow direct uncertainty quantification for quantities of interest such as hazard ratios $\exp(\beta)$, median survival times, survival curves at selected covariate values, and contrasts between treatments or groups. -->

## Survival data

Our data are bacterial abundance profiles from stool samples of laboratory grown mice [FIX: add ref]. The data includes measurements from 150 mice and 48 microbial genera, with follow-up recorded as the time to type 1 diabetes onset or censoring at [FIX] days. Approximately half of the mice developed type diabetes during the follow-up. 

The data is included in the R package $mia$ which contains tools for microbiome data analysis, and we will load the data from there. 

All continuous microbial abundances (log transformed proportions) are standardized to have mean 0 and standard deviation 1 before modeling. Consequently, hazard ratios are interpreted per one standard deviation increase in abundance. 

```{r, message=FALSE, warning=FALSE}
# Packages
library(tidyverse)
library(rstan)
library(posterior) # Where needed? [FIX?]
library(survival)
library(TreeSummarizedExperiment)
library(survminer)
library(mia)
```

```{r, warning=FALSE, message=FALSE}
# Load TSE
tse <-  readRDS(file.path("data", "survival_tse.rds"))

# Relative abundance + log transform
tse <- tse |>
  transformAssay(method = "relabundance", assay.type = "counts", name = "relabundance") |>
  transformAssay(method = "log", assay.type = "relabundance", pseudocount = 1e-6, name = "log_abund")

# Features (samples x microbes)
df <- as.data.frame(t(assay(tse, "log_abund")))

# Outcomes
df$Event      <- colData(tse)$Event
df$Event_time <- colData(tse)$Event_time

# Pick 10 random microbes as covariates
# [FIX: pick covariates that produce at least some non-zero effects]
set.seed(2025)
feature_cols <- setdiff(names(df), c("Event", "Event_time"))
selected_microbes <- sample(feature_cols, min(10L, length(feature_cols)))

# Build design matrix (scaled)
X <- scale(as.matrix(df[, selected_microbes, drop = FALSE]))
K <- ncol(X)

# Analysis data frame
analysis_df <- data.frame(
  time  = as.numeric(df$Event_time),
  event = as.integer(df$Event),
  X,
  check.names = FALSE
)
rownames(analysis_df) <- rownames(df)

# Stan data
standata <- list(
  N    = nrow(analysis_df),
  K    = K,
  time = as.vector(analysis_df$time),
  event= as.integer(analysis_df$event),
  X    = as.matrix(analysis_df[, colnames(X), drop = FALSE])
)

```

## Weibull survival model

The Weibull model is a versatile parametric choice for time-to-event data that uses the Weibull distribution for the event times $T$. The distribution has two parameters: shape $k>0$ and scale  $\lambda>0$. A common way to include parameters in the model is to set the exponentiated linear term as the scale parameter: 

$$\lambda = \exp(\beta_0 + \beta_1X_1 + \ldots + \beta_KX_K),$$where $X_i$ are the covariate values and $\beta_i$ the corresponding regression coefficients. The scale parameter calibrates the time axis, with larger values implying longer survival. 

The shape parameter, on the other hand, controls how the risk evolves over time: when $0<k<1$ the hazard decreases, with $k=1$ it is constant, and when $k>1$ it increases. This ability to capture decreasing, constant, or increasing hazards with a single parameter, makes the Weibull distribution is a good starting point for event time modeling.


The survival and hazard functions can be shown to take the following form [FIX: add reference]: 

$$
S(t)=\exp\left[-\left(\frac{t}{\lambda}\right)^{k}\right],
$$ 
and
$$ 
h(t) = \frac{k}{\lambda}\left(\frac{t}{\lambda}\right)^{k-1}. 
$$
A convenient summary is the median survival time 
$$ 
\operatorname{med}(T)=\lambda(\log 2)^{1/k}, 
$$
which increases linearly with $\lambda$ and, for a fixed $\lambda$, decreases as $k$ increases.

It is easy to show that increasing one covariate $x_j$ by one unit, while holding other covariates constant results in an increase

$$
\mathrm{HR}_j \;=\; \frac{h\!\left(t \mid x_j + 1\right)}{h\!\left(t \mid x_j\right)} \;=\; \exp\!\left(-k\,\beta_j\right).
$$

Thus a positive $\beta_j$ yields $\mathrm{HR}_j<1$ (lower hazard; protective) and a negative $\beta_j$ yields $\mathrm{HR}_j>1$ (higher hazard). 



### Example

Next, we illustrate the Weibull survival model by estimating effects and the corresponding hazard ratios of bacterial genera on diabetes incidence. 

Stan program for the model is below. The data likelihood is coded using the `target +=` notation. This simply means that the term after this notation is added in log likelihood of the model, and is equivalent to the standard `~` notation. [FIX: why use the target notation?] The program is implemented to also accommodate covariate-free regression, thus the zero lower bound for the number of covariates `K`.

<!-- Kaplan-Meier -->
<!-- A simple descriptive summary is the Kaplan–Meier curve, which estimates the survival function $$ -->
<!-- S(t) = P(T > t) -->
<!-- $$and handles right censoring naturally. We will use Kaplan–Meier estimates for exploratory checks alongside model-based summaries. -->


```{stan output.var="weibull_model"}
// Weibull survival model (shape k, scale λ), simple AFT form
data {
  int<lower=1> N;                    // number of observations
  int<lower=0> K;                    // number of covariates
  vector<lower=0>[N] time;           // observed times t_i
  int<lower=0,upper=1> event[N];     // 1=event, 0=right-censored
  matrix[N, K] X;                    // covariates (N x K, K can be 0)
}

parameters {
  real<lower=0> k;                   // Weibull shape
  real beta0;                        // intercept
  vector[K] beta;                    // coefficients
}

transformed parameters {
  vector<lower=0>[N] lambda;
  
  if (K > 0) lambda = exp(beta0 + X * beta);
    else lambda = exp(rep_vector(beta0, N));
}

model {
  // weakly-informative priors
  k     ~ lognormal(0, 1);
  beta0 ~ normal(0, 1);
  if (K > 0) beta ~ normal(0, 1);

  // likelihood
  // [FIX: factorize, without for loop]
  for (i in 1:N) {
    if (event[i] == 1) // events
      target += weibull_lpdf(time[i] | k, lambda[i]);
    else // censored cases
      target += weibull_lccdf(time[i] | k, lambda[i]);  // log S(t_i)
  }
}

generated quantities {
  vector[N] t_median; // median survival time
  for (i in 1:N)
    t_median[i] = lambda[i] * pow(log(2), 1.0 / k);
}

```



Next, we fit the model and obtain the HRs by transforming the posterior draws of $k$ and $\beta_j$ with $\exp(-k\,\beta_j)$, and then use the posterior median and the $95%$ CIs and our summaries. We visualize the results with an ordered forest plot highlighting covariates whose CI excludes 1.

[FIX: check code]

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit
fit <- rstan::sampling(
  object  = weibull_model,
  data    = standata,
  seed    = 2025, refresh = 0
)

# Print summary
# print(fit, pars = c("k","beta0","beta"))

# Posterior draws
M <- posterior::as_draws_matrix(fit)
k_draw <- M[, "k"]
K <- standata$K
covar_names <- colnames(standata$X)

# Hazard ratios
hr_mat <- do.call(rbind, lapply(seq_len(K), function(j) {
  bj <- M[, paste0("beta[", j, "]")]
  hr_draw <- exp(- k_draw * bj)
  c(
    med = median(hr_draw),
    lo  = unname(quantile(hr_draw, 0.025)),
    hi  = unname(quantile(hr_draw, 0.975))
  )
}))
hr_df <- as.data.frame(hr_mat)
hr_df$term <- covar_names
names(hr_df) <- c("HR","lower","upper","term")

# Significance
hr_df$signif <- (hr_df$lower > 1) | (hr_df$upper < 1)
hr_df <- hr_df[order(hr_df$HR), ]
hr_df$term <- factor(hr_df$term, levels = hr_df$term)

# Plot a forest plot of hazard ratios
ggplot(hr_df, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"), guide = "none") +
  labs(title = "Weibull hazard ratios",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

As evident in the the plot, several taxa exclude 1 implying posterior evidence for an association between the genus and diabetes. 


To assess *discrimination*, that is, how well the model is able to correctly rank cases bases on event risk, we form a risk score from the linear predictors $\log\lambda_i=\beta_0+x_i^\top\beta$. We quantify discrimination with Harrell's C-index, which is the ratio of correctly ranked cases and all comparable pairs. Comparability is determined by censoring; two cases cannot be compared if event times for both are censored, or if one case is censored before the other experiences an event. 
[FIX: compute and compare C posteriors]

We can also visualize discrimination ability by using posterior medians of $\beta_0$ and $\beta$. By splitting subjects at the median score value (Low vs High), we can generate the respective Kaplan–Meier curves:

[FIX: remove p-value; K-M plot with median score split per posterior sample?]

```{r, fig.width=10, fig.height=6, dpi=300, warning=FALSE}
# Posterior medians for the AFT linear predictor
beta0_med <- median(M[, "beta0"])
beta_med  <- sapply(seq_len(K), function(j) median(M[, paste0("beta[", j, "]")]))

# Model-based score
risk_lp <- as.numeric(beta0_med + as.matrix(standata$X) %*% beta_med)

# Data for KM
dat_km <- data.frame(
  time  = standata$time,
  event = standata$event,           # 1 = event, 0 = censored
  risk  = risk_lp
)

# Median split (High score = better survival)
cut_pt <- median(dat_km$risk, na.rm = TRUE)
dat_km$group <- factor(
  ifelse(dat_km$risk > cut_pt, "High score", "Low score"),
  levels = c("Low score", "High score")
)

# Kaplan–Meier fit and plot
sf <- survfit(Surv(time, event) ~ group, data = dat_km)

ggsurvplot(
  sf, data = dat_km,
  conf.int    = TRUE,
  pval        = TRUE,
  legend.labs = levels(dat_km$group),
  palette     = c("#E64B35", "#4DBBD5"),
  xlab = "Time", ylab = "Survival probability"
)

```

Kaplan–Meier curves by the model-based score (median split) show clear separation: the high AFT score group (larger $\log\lambda$, implying longer times) has higher survival (curve above), while the low-score group declines faster.

<!-- The log-rank test indicates a statistically significant difference. -->


<!-- Here -->


:::::::::::::::::::::::::::::: challenge
Generate a posterior predictive distribution for the event times. 
::::::::::::::::::::::::::::::::::::::::

<!-- :::::::::::::::::::::::::::::: challenge -->

<!-- Pick two microbial covariates from the Weibull forest plot:   -->
<!-- one with HR < 1 and one with HR > 1. -->

<!-- 1. Explain in words what each hazard ratio means for time to T1D onset.   -->
<!-- 2. Remember that covariates are standardized: interpret the HRs per one–standard deviation increase. -->


<!-- ::::::::::::::::::::::: solution -->

<!-- **HR < 1**   -->
<!-- A one–standard deviation increase in the microbial covariate is associated with a lower instantaneous risk of T1D onset (protective effect). -->

<!-- **HR > 1**   -->
<!-- A one–standard deviation increase in the microbial covariate is associated with a higher instantaneous risk of T1D onset (harmful effect). -->

<!-- :::::::::::::::::::::::::::::::: -->

<!-- :::::::::::::::::::::::::::::::::::::::: -->



## Cox proportional hazards model

The Cox model is a semiparametric approach for time-to-event data in which covariates act multiplicatively on the hazard. Let $T$ be the event time and $x$ the covariates. The hazard is

$$
h(t \mid x) \;=\; h_0(t)\,\exp\!\bigl(x^\top \beta\bigr),
$$

where $h_0(t)$ is an unspecified baseline hazard. Through the cumulative baseline hazard $H_0(t)=\int_0^t h_0(u)\,du$, the survival function can be written

$$
S(t \mid x) \;=\; \exp\!\Bigl[-\,H_0(t)\,\exp\!\bigl(x^\top \beta\bigr)\Bigr]
\;=\; S_0(t)^{\,\exp(x^\top \beta)}.
$$

The coefficients $\beta$ are log–hazard ratios: a one-unit increase in $x_j$ multiplies the hazard by $\exp(\beta_j)$. Estimation is based on the partial likelihood, which compares each event’s linear predictor to those of individuals still at risk at that time and avoids modeling $h_0(t)$. The key assumption is proportional hazards, meaning the hazard ratio for two covariate values is constant over time.

::::::::::::::::::::: callout
Baseline hazard can be estimated using splines, which enables survival time predictions. 

In brms this can be done as follows: [FIX]
:::::::::::::::::::::::::::::

### Example

Next, we illustrate the Cox proportional hazards model by estimating hazard ratios and comparing risk groups with Kaplan–Meier curves. We fit the model via the partial likelihood (leaving the baseline hazard unspecified) and then summarize covariate effects as exponentiated coefficients $exp(\beta)$. Below is the structure of the Stan program.

```{stan output.var="cox_model"}
// Cox proportional hazards model, partial likelihood (PH form)
data {
  int<lower=1> N;                    // observations
  int<lower=0> K;                    // number of covariates
  vector<lower=0>[N] time;           // observed times t_i
  int<lower=0, upper=1> event[N];    // 1=event, 0=right-censored
  matrix[N, K] X;                    // covariates (N x K, K can be 0)
}

transformed data {
  // Indices of observed events 
  int<lower=0> n_event = 0;
  int event_idx[N];
  for (i in 1:N) {
    if (event[i] == 1) {
      n_event += 1;
      event_idx[n_event] = i;
    }
  }
}

parameters {
  vector[K] beta;                    // log hazard ratios
}

model {
  // Weakly-informative prior
  if (K > 0) beta ~ normal(0, 1);

  // Cox partial log-likelihood
  vector[N] eta;
  if (K > 0) eta = X * beta; else eta = rep_vector(0, N);

  for (e in 1:n_event) {
    int i = event_idx[e];
    real lse = negative_infinity();   // log(0)
    for (j in 1:N) {
      if (time[j] >= time[i])         // j in risk set at t_i
      lse = log_sum_exp(lse, eta[j]);
    }
    target += eta[i] - lse;
  }
}

generated quantities {
  vector[K] HR;                        // per-covariate hazard ratios
  for (k in 1:K) HR[k] = exp(beta[k]);
}

```

In our Cox proportional hazards setup the hazard for subject $i$ is modeled as $$
h_i(t) = h_0(t)\,\exp\!\big(x_i^\top\beta\big),
$$ where $h_0(t)$ is an unspecified baseline hazard and $\beta$ are log–hazard ratio coefficients. If we increase a single covariate $x_j$ by one unit (holding the others fixed), the hazard scales by $$
\mathrm{HR}_j \;=\; \frac{h\!\left(t \mid x_j + 1\right)}{h\!\left(t \mid x_j\right)} \;=\; \exp(\beta_j).
$$ Unlike the Weibull AFT parameterization (where $\mathrm{HR}_j=\exp(-k\,\beta_j)$ so a positive $\beta_j$ is protective), in the Cox PH model a positive $\beta_j$ yields $\mathrm{HR}_j>1$ (higher hazard) and a negative $\beta_j$ yields $\mathrm{HR}_j<1$ (lower hazard is protective). In the code, we obtain HRs directly from posterior draws of $\beta_j$ via $\exp(\beta_j)$ and summarize them by the posterior median and the $95\%$ credible interval (2.5%–97.5%).

The next chunk transforms posterior draws of $\beta_j$ into hazard-ratio draws via $\exp(\beta_j)$, summarizes each covariate by the posterior median and 95% credible interval, and then produces an ordered forest plot highlighting covariates whose intervals do not include 1. The model is fit with 4 chains and 4000 iterations per chain.

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit the Cox model
fit <- rstan::sampling(
  object  = cox_model,
  data    = standata,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

# Print summary
print(fit, pars = c("beta"))

# Posterior draws
M <- posterior::as_draws_matrix(fit)
K <- standata$K
covar_names <- colnames(standata$X)

# Cox PH hazard ratios
hr_mat <- do.call(rbind, lapply(seq_len(K), function(j) {
  bj <- M[, paste0("beta[", j, "]")]
  hr_draw <- exp(bj)
  c(
    med = median(hr_draw),
    lo  = unname(quantile(hr_draw, 0.025)),
    hi  = unname(quantile(hr_draw, 0.975))
  )
}))
hr_df <- as.data.frame(hr_mat)
hr_df$term <- covar_names
names(hr_df) <- c("HR","lower","upper","term")

# Significance
hr_df$signif <- (hr_df$lower > 1) | (hr_df$upper < 1)
hr_df <- hr_df[order(hr_df$HR), ]
hr_df$term <- factor(hr_df$term, levels = hr_df$term)

# Plot a forest plot of hazard ratios
ggplot(hr_df, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"), guide = "none") +
  labs(title = "Cox PH hazard ratios",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

For the Cox PH model, the hazard‐ratio pattern mirrors the Weibull results: hazard ratios left of 1 (with 95% intervals not crossing 1) suggest protective associations, while those right of 1 indicate increased hazard; intervals that cross 1 are inconclusive. In our analysis, the Cox HRs align closely with the Weibull HRs, leading to very similar forest plots.

To visualize how the Cox model separates risk, we form a model-based risk score from the PH linear predictor $\eta_i = x_i^\top \beta$ using posterior medians of $\beta$. We split subjects at the median of this score (“Low” vs “High”) and compare their Kaplan–Meier curves with a log-rank test. The Kaplan–Meier estimator is nonparametric and does not depend on the Cox model itself, only the grouping into “Low” and “High” risk comes from the model, so the same KM machinery could equally be used with the Weibull-based risk score.

```{r, fig.width=10, fig.height=6, dpi=300, warning=FALSE}
# Posterior medians for the PH linear predictor
beta_med <- sapply(seq_len(K), function(j) median(M[, paste0("beta[", j, "]")]))
risk_lp  <- as.numeric(as.matrix(standata$X) %*% beta_med)

dat_km <- data.frame(
  time  = standata$time,
  event = standata$event,
  risk  = risk_lp
)

# Median split of the risk score (High risk = worse survival)
cut_pt <- median(dat_km$risk, na.rm = TRUE)
dat_km$group <- factor(
  ifelse(dat_km$risk > cut_pt, "High risk", "Low risk"),
  levels = c("High risk", "Low risk")
)

# Kaplan–Meier fit and plot
sf <- survfit(Surv(time, event) ~ group, data = dat_km)

ggsurvplot(
  sf, data = dat_km,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Model-based risk",
  legend.labs  = levels(dat_km$group),
  palette      = c("#E64B35", "#4DBBD5"),
  xlab         = "Time", ylab = "Survival probability"
)


```

Kaplan–Meier curves by the model-based score (median split) show clear separation: the high PH risk group (larger $\eta = x^\top\beta$, implying higher hazard) has lower survival (curve below), while the low-risk group remains higher. The log-rank test indicates a statistically significant difference.

## Hierarchical survival model

In the final example we switch from the microbiome data to a clinical survival dataset. The cancer data from the survival package contains follow-up on patients with advanced lung cancer, with time recorded from study entry to death or censoring. Each patient has several baseline covariates, including age, sex, ECOG performance status, physician- and patient-rated Karnofsky scores, daily caloric intake, and weight loss in last six months. In addition, patients are treated at different institutions, indexed by the variable inst.

To capture both individual-level risk factors and between-institution heterogeneity, we construct a hierarchical Cox proportional hazards model. All baseline covariates enter the linear predictor as fixed effects, while the institution index defines group-specific frailties that are modeled as random effects on the log-hazard scale. For patient $i$ treated at institution $g[i]$, the hazard is

$$
h_i(t) \;=\; h_0(t)\,\exp\bigl(x_i^\top \beta + u_{g[i]}\bigr), 
\qquad u_g \sim \mathcal{N}(0, \sigma_u^2),
$$

where $x_i$ is the vector of scaled baseline covariates, $\beta$ are fixed-effect coefficients, and $u_g$ is the institution-specific frailty term with variance $\sigma_u^2$. All baseline covariates are standardized (mean 0, standard deviation 1), so hazard ratios summarize the effect of a one–standard deviation increase in each covariate.

```{r}
# Data
data(cancer, package = "survival")

# Complete cases
cancer2 <- na.omit(cancer)

# Outcome: time and event indicator (status: 1=censored, 2=dead)
cancer2$time  <- cancer2$time
cancer2$event <- as.integer(cancer2$status == 2)

# Grouping factor: institution
cancer2$group <- factor(cancer2$inst)
G <- nlevels(cancer2$group)

# All baseline covariates
X_mm <- model.matrix(
  ~ age + sex + ph.ecog + ph.karno + pat.karno + meal.cal + wt.loss,
  data = cancer2
)[, -1, drop = FALSE]  # drop intercept

# Scale columns
X <- scale(X_mm)
K <- ncol(X)

# Analysis data frame
analysis_df_cancer <- data.frame(
  time  = cancer2$time,
  event = cancer2$event,
  group = cancer2$group,
  X,
  check.names = FALSE
)

# Stan data list for hierarchical Cox PH
standata_cancer <- list(
  N     = nrow(analysis_df_cancer),
  G     = G,
  group = as.integer(analysis_df_cancer$group),
  K     = K,
  time  = as.vector(analysis_df_cancer$time),
  event = as.integer(analysis_df_cancer$event),
  X     = as.matrix(analysis_df_cancer[, colnames(X), drop = FALSE])
)

```

Next we implement this hierarchical specification in Stan by extending the Cox proportional hazards model with a random intercept for each institution. The linear predictor combines fixed effects for the scaled covariates with a group-specific frailty term, and inference proceeds again via the partial likelihood. As before, covariate effects are expressed as hazard ratios obtained by exponentiating the regression coefficients, allowing us to quantify how each predictor multiplies the event hazard while the frailty term accounts for unexplained heterogeneity between institutions.

```{stan output.var="cox_hier_model"}
// Stan: hierarchical Cox proportional hazards model
data {
  int<lower=1> N;                     // observations
  int<lower=1> G;                     // groups
  int<lower=1,upper=G> group[N];      // group index
  int<lower=0> K;                     // number of covariates
  vector<lower=0>[N] time;            // observed times t_i
  int<lower=0,upper=1> event[N];      // 1=event, 0=censored
  matrix[N, K] X;                     // covariates
}

transformed data {
  int<lower=0> n_event = 0;
  int event_idx[N];
  for (i in 1:N) {
    if (event[i] == 1) {
      n_event += 1;
      event_idx[n_event] = i;
    }
  }
}

parameters {
  vector[K] beta;                     // log hazard ratios
  vector[G] u_raw;                    // group effects
  real<lower=0> sigma_u;              // group SD
}

model {
  vector[N] eta;

  if (K > 0) beta ~ normal(0, 1);
  u_raw   ~ normal(0, 1);
  sigma_u ~ normal(0, 1);

  for (i in 1:N)
    eta[i] = (K > 0 ? X[i] * beta : 0) + sigma_u * u_raw[group[i]];

  for (e in 1:n_event) {
    int i = event_idx[e];
    real lse = negative_infinity();
    for (j in 1:N) {
      if (time[j] >= time[i])
        lse = log_sum_exp(lse, eta[j]);
    }
    target += eta[i] - lse;
  }
}

generated quantities {
  vector[K] HR;                       // hazard ratios
  real var_frailty = square(sigma_u); // between-group variance

  for (k in 1:K)
    HR[k] = exp(beta[k]);
}


```

```{r, fig.width=10, fig.height=6, dpi=300}
# Fit hierarchical Cox and plot Hazard Ratios
fit_cancer_cox_hier <- rstan::sampling(
  object  = cox_hier_model,
  data    = standata_cancer,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

# Summary: fixed effects + frailty variance
print(fit_cancer_cox_hier, pars = c("beta","sigma_u","var_frailty","HR"))

# Posterior draws
M_cancer_cox <- posterior::as_draws_matrix(fit_cancer_cox_hier)
K <- standata_cancer$K
covar_names_cancer <- colnames(standata_cancer$X)

# Cox PH hazard ratios
hr_df_cancer <- do.call(rbind, lapply(seq_len(K), function(j) {
  hr_draw <- exp(M_cancer_cox[, paste0("beta[", j, "]")])
  data.frame(
    HR    = median(hr_draw),
    lower = unname(quantile(hr_draw, 0.025)),
    upper = unname(quantile(hr_draw, 0.975)),
    term  = covar_names_cancer[j],
    stringsAsFactors = FALSE
  )
}))

hr_df_cancer$signif <- (hr_df_cancer$lower > 1) | (hr_df_cancer$upper < 1)
hr_df_cancer <- hr_df_cancer[order(hr_df_cancer$HR), , drop = FALSE]
hr_df_cancer$term <- factor(hr_df_cancer$term, levels = hr_df_cancer$term)

ggplot(hr_df_cancer, aes(y = term, x = HR, color = signif)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey70") +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.6) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "#2C3E50"),
                     guide = "none") +
  labs(title = "Hierarchical Cox PH hazard ratios (cancer data)",
       x = "Hazard ratio (log scale)", y = NULL) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

Performance status variables (ph.ecog and ph.karno) have hazard ratios above 1, with intervals not overlapping 1, indicating strong evidence that poorer performance status is associated with higher mortality. Age also shows a harmful effect, though with greater posterior uncertainty. In contrast, patient-rated Karnofsky score and caloric intake (meal.cal) have hazard ratios centered below or near 1, suggesting weak or mild protective associations that are mostly uncertain. Weight loss shows a hazard ratio slightly below/around 1, but with wide uncertainty, so its association with survival is unclear. Sex (female lower hazard) appears protective with a hazard ratio below 1.

The frailty variance is small but non-zero, indicating modest but detectable between-institution heterogeneity: some institutions have slightly higher or lower baseline hazard even after adjusting for patient-level predictors, but these differences are limited in magnitude compared to fixed effects.

Overall, performance status dominates the risk profile, while other baseline variables show smaller and more uncertain associations. The hierarchical model accounts for institutional heterogeneity but the main structure of covariate effects remains driven by individual-level predictors.

Next we split subjects at the median of this score (“Low” vs “High”) and compare their Kaplan–Meier curves with a log-rank test.

```{r, fig.width=10, fig.height=6, dpi=300}
# Posterior medians for the PH linear predictor
beta_med_cancer <- sapply(seq_len(K), function(j) {
  median(M_cancer_cox[, paste0("beta[", j, "]")])
})

# Model-based risk score (higher = worse survival)
risk_lp_cancer <- as.numeric(as.matrix(standata_cancer$X) %*% beta_med_cancer)

dat_km_cancer <- data.frame(
  time  = standata_cancer$time,
  event = standata_cancer$event,
  risk  = risk_lp_cancer
)

# Median split of the risk score (High risk = worse survival)
cut_pt_cancer <- median(dat_km_cancer$risk, na.rm = TRUE)
dat_km_cancer$group <- factor(
  ifelse(dat_km_cancer$risk > cut_pt_cancer, "High risk", "Low risk"),
  levels = c("High risk", "Low risk")
)

# Kaplan–Meier fit and plot
sf_cancer <- survfit(Surv(time, event) ~ group, data = dat_km_cancer)

ggsurvplot(
  sf_cancer, data = dat_km_cancer,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Model-based risk",
  legend.labs  = levels(dat_km_cancer$group),
  palette      = c("#E64B35", "#4DBBD5"),
  xlab         = "Time", ylab = "Survival probability"
)

```

Kaplan–Meier curves by the model-based score (median split) show clear separation: the high PH risk group has lower survival, while the low-risk group remains higher. The log-rank test indicates a statistically significant difference.

::: keypoints
-   Survival analysis models time until an event while handling censoring and aims to describe and compare time-to-event patterns across covariates or groups.

-   Time-to-event outcomes can be modeled parametrically with a Weibull model that specifies survival and hazard functions or semiparametrically with a Cox proportional hazards model that leaves the baseline hazard unspecified while estimating relative risk.

-   The hazard summarizes instantaneous event risk, hazard ratios compare risk between groups or covariate levels, and Kaplan–Meier curves provide a nonparametric survival estimate often shown with a log-rank test.
:::



:::::::::::::::::::::::::::::: challenge

Fit a Bayesian Cox proportional hazards model with sex as the only covariate and then compare the model-based hazard ratio with nonparametric Kaplan–Meier curves.

1. Fit the Cox PH model in Stan using the code above and inspect the posterior summary for `beta` and `HR`.  
2. Compute the posterior median hazard ratio (`HR_med`) and a 95% credible interval (`HR_ci`) for Female vs Male and interpret these in words.  
3. Plot Kaplan–Meier curves for males and females and compare them to the Cox model’s posterior hazard ratio: do they tell a consistent story about differences in survival by sex?


::::::::::::::::::::::: solution

```{r}
# Data
data(cancer, package = "survival")
cancer2 <- na.omit(cancer)

# Outcome: time and event (status: 1=censored, 2=dead)
time  <- cancer2$time
event <- as.integer(cancer2$status == 2)

# Covariate: sex (0 = Male, 1 = Female)
sex_num <- ifelse(cancer2$sex == 2, 1, 0)

# X matrix
X_sex <- matrix(sex_num, ncol = 1)
colnames(X_sex) <- "sex_female"
K <- ncol(X_sex)

standata_sex <- list(
  N    = length(time),
  K    = K,
  time = as.vector(time),
  event= as.integer(event),
  X    = X_sex
)


# Fit Cox PH -model
fit_cox_sex <- rstan::sampling(
  object  = cox_model,      # same as in the lecture
  data    = standata_sex,
  chains  = 4, iter = 4000, warmup = 1000,
  seed    = 2025, refresh = 0,
  control = list(adapt_delta = 0.95)
)

```


```{r}
print(fit_cox_sex, pars = c("beta", "HR"))
```


```{r}
M_sex <- posterior::as_draws_matrix(fit_cox_sex)

# log-hazard ratio Female vs Male
beta_draw  <- M_sex[, "beta[1]"]
HR_draw    <- exp(beta_draw)

HR_med     <- median(HR_draw)
HR_ci      <- quantile(HR_draw, c(0.025, 0.975))
HR_med
HR_ci

```


```{r}
# Kaplan-Meier
cancer2$sex_factor <- factor(cancer2$sex, levels = c(1, 2),
                             labels = c("Male", "Female"))

sf_sex <- survfit(Surv(time, event) ~ sex_factor, data = cancer2)

ggsurvplot(
  sf_sex, data = cancer2,
  conf.int     = TRUE,
  pval         = TRUE,
  legend.title = "Sex",
  legend.labs  = c("Male", "Female"),
  xlab         = "Time", ylab = "Survival probability"
)

```



1. Posterior median hazard ratio (Female vs Male) is about **0.63**. Females have roughly **37% lower** instantaneous risk of death than males at any given time.

2. The 95% credible interval is about **(0.43, 0.91)**, entirely below 1. This provides clear evidence that females have a lower hazard of death than males in this dataset.

3. The Kaplan–Meier curve for females lies above the curve for males. Nonparametric KM estimates also indicate better survival for women, consistent with the Cox model’s hazard ratio.
   
::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::


## Resources

-   User’s guide <https://mc-stan.org/docs/stan-users-guide/survival.html>
